---
title: 'STOR 565: IA Solution For Homework 6'
author: "Brian N. White"
date: "2/23/2022"
output:
  html_document: default
  pdf_document: default
---
```{r}
# set global code-chunk options
knitr::opts_chunk$set(fig.align = 'center')
```

```{r, include = F}
library(tidyverse)
library(reshape2)
library(png)
library(grid)
```

```{r}
# import data
nba <- read.csv('./data/nba-teams-2017.csv')
```

**1 (a)**

Create box plots of the quantitative features (i.e. all but) teams to see if you should scale the data when performing PCA. Describe your findings in words.

The variance of the features other than 'wins' and 'points' are comparable but differ greatly from the variance of 'wins' and 'points'. The data should be scaled.
```{r, warning = F, message = F}
nba <- nba %>%
  select(team, wins, points, points3, free_throws, off_rebounds, def_rebounds, assists, steals, personal_fouls)

# box plots for un-scaled features
nba %>%
  select(-team) %>%
  melt() %>%
  ggplot(aes(y = value, color = variable)) +
  geom_boxplot(outlier.size = 0.8)

# box plots for scaled features
nba %>%
  select(-team) %>%
  scale(center = T) %>% 
  as.tibble() %>%
  melt() %>%
  ggplot(aes(y = value, color = variable)) +
  geom_boxplot(outlier.size = 0.8)

# check that data are, in fact, centered and scaled
nba %>%
  select(-team) %>%
  scale() %>%
  as.tibble() %>%
  summarise_all(.funs = list(sd = sd, mean = mean))

# using base R
boxplot(scale(nba[,-1], center = T))
```
**1 (b)**

Obtain PC loadings of the first four princple components (PCs). Only display the first few elements of each loading in your report.

```{r}
# compute the principle components
nba_pc <- nba %>%
  select(-team) %>%
  scale(center = T) %>%
  prcomp()

# loadings
nba_pc$rotation
# general summary
summary(nba_pc)
```
```{r}
# direct computation
nba_svd <- nba %>%
  select(-team) %>%
  scale() %>%
  svd()

# standard deviations of principle components 
nba_pc$x %>%
  as.tibble() %>%
  summarise_all(sd)

nba_pc$sdev

# compare the principle components
as.matrix(scale(nba[, -1]), center = T )%*% nba_svd$v[,2]
nba_pc$x[,2]

# compare proportion of variance explained
nba_svd$d^2/sum(nba_svd$d^2)
summary(nba_pc)$importance[2,]

# compare loadings
nba_svd$v
nba_pc$rotation
```

**1 (c)**

Generate a scree plot describing the amount explained by the various PCs.

```{r}
# contains the variance of proportion explained for each principle component
prop_var <- summary(nba_pc)$importance[2,]

# create the scree plot
data.frame(x = 1:9, y = prop_var) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_line(col = 'light blue') +
  scale_x_continuous(breaks = 1:length(prop_var)) +
  labs(x = 'Principle component', y = 'Proportion of variance', title = 'Scree Plot')
```

**1 (d)**

Make another plot showing the cumulative percent of the variance explained.

```{r}
# contains the cumulative variance explained with each additional pc added
cum_var <- summary(nba_pc)$importance[3,]

# create the scree plot
data.frame(x = 1:length(cum_var), y = cum_var) %>%
  ggplot() +
  geom_point(aes(x = x, y = y)) +
  geom_text(aes(x =x , y = y, label =scales::percent(cum_var)), nudge_x = 0.25, nudge_y = 0.04) +
  geom_line(aes(x = x, y = y), col = 'light blue') +
  scale_x_continuous(breaks = 1:length(cum_var)) +
  labs(x = 'No. of principle components', y = 'Cumulative variance', title = 'Cumulative Variance Explained Plot')
```

**1 (e)**

If you were to retain all PCs which explain at least 90% of the variance, how many PCs would you retain?

One should retain the first six principle components to capture at least 90% of the variance while maximizing the decrease in dimension. 

```{r}
# confirm results 
which(summary(nba_pc)$importance[3,] >= 0.9)
# cumulative proportion of variance explained by the 6th pc
summary(nba_pc)$importance[3, 6]
```


**1 (d)**

Plot PC1 vs PC2 with the team names and try to interpret your findings.

```{r}
# create data frame of pc scores
df <- data.frame(team = nba$team, nba_pc$x) 
# assign row names to df for use in plot
rownames(df) <- df$team

# create plot of pc1 vs pc2
df %>%
  ggplot(aes(x = PC1, y = PC2, label = rownames(df)))  +
  geom_text(check_overlap = T, size = 3) +
  xlim(-6, 3)
```

**2 (a)**

```{r}
# import redfin png
redfin <- readPNG('./data/Redfin_house.png')[, , 1]

# dimensions of data frame
dim(redfin)

# plot image
grid.raster(redfin)

# examine histogram of pixel intensities
data.frame(x = as.vector(redfin)) %>%
  ggplot(aes(x = x)) + 
  geom_histogram() +
  labs(title = 'Histogram of pixel intensities')
```

**2 (b)**

```{r}
redfin_pc <- redfin %>% 
                  prcomp(center = F, scale = F)

redfin_prop_var <- summary(redfin_pc)$importance[2,]
redfin_cum_var <- summary(redfin_pc)$importance[3,]

data.frame(x = 1:length(redfin_prop_var), y = redfin_prop_var) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_line(col = 'light blue') +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = seq(0, length(redfin_prop_var), by = 100)) +
  labs(x = 'Principle component', y = 'Proportion of variance', title = 'Scree Plot')

# cumulative variance plot
data.frame(x = 1:length(redfin_prop_var), y = redfin_cum_var) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_line(col = 'light blue') +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = seq(0, length(redfin_cum_var), by = 100)) +
  labs(x = 'No. of principle components', y = 'Cumulative variance', title = 'Cumulative Variance Explained Plot')

# check minimum no. of principle components needed to explain 90% of the variation
which(summary(redfin_pc)$importance[3,] > 0.9)[1]
```

**2 (c)**

```{r, cache = T}
# compute SVD of redfin
redfin_svd <- svd(redfin)

# compute compressed redfin for different numbers of included principle components
for(k in c(1, 5, 10, 15, 20, 30, 50, 100, 200)){
  
  if(k == 1) {
    
    redfin_proj <- redfin_svd$d[1]*as.matrix(redfin_svd$u[, 1:k]) %*% as.matrix(t(redfin_svd$v[, 1:k]))
    
  } else {
  
redfin_proj <- as.matrix(redfin_svd$u[, 1:k]) %*% as.matrix(diag(redfin_svd$d[1:k])) %*% as.matrix(t(redfin_svd$v[, 1:k]))

}

grid.newpage(recording = F)
grid.raster(pmax(pmin(redfin_proj, 1), 0))

}
```

